### 2019-12-10 Statistical Methods/02_numerical-summaries.rmd
statistical methods, ch2
statistical methods, ch2

@@

@@

@@

```{r, echo = T}
```

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

```{r, echo = T}
x <- 1:10

sum(x^2)

(sum(x))^2
```

@@

```{r, echo = T}
x <- c(-1, 0, 3, 0, 2, -5)

mean(x)
median(x)
```

@@

```{r, echo = T}
x <- c(2, 2, 3, 10, 100, 1000)

mean(x)
median(x)
```

@@

```{r, echo = T}
scores <- c(73, 74, 92, 98, 100, 72, 74, 85, 76, 94, 89, 73, 76, 99)

mean(scores)
mean(scores, trim = .2)
median(scores)
```

@@

```{r, echo = T}
23 * 14.7
```

@@

```{r, echo = T}
x <- c(3, 6, 8, 12, 23, 26, 37, 42, 49, 63)
mean(x)
```

@@

```{r, echo=T}
x <- c(3, 6, 8, 12, 23, 26, 37, 42, 49, 100)
mean(x)
```

@@

```{r, echo=T}
x <- c(3, 6, 8, 12, 23, 26, 37, 42, 49, 1000)
mean(x)
```

@@

_The mean is very sensitive to outliers._

@@

```{r, echo=T}
x <- c(3, 6, 8, 12, 23, 26, 37, 42, 49, 100)
median(x)
```

```{r, echo=T}
x <- c(3, 6, 8, 12, 23, 26, 37, 42, 49, 1000)
median(x)
```

@@

_One._

@@

mean = $g = (.2N), g + 1$

median = ~$.5N$

### 2.11

For the values 0, 23, -1, 12, -10, -7, 1, -19, -6, 12, 1, -3 compute the lower and upper quartiles using the ideal fourths.

```{r, echo = T}
# Used in 11 & 12
idealf <- function( x ) {
   n <- length(x)
   sorted <- sort(x)
   i <- (n / 4) + 5/12
   j <- floor(i)
   h <- i - j
   q1 <- (1 - h)*sorted[j] + h*sorted[j + 1]

   k <- n - j + 1
   q2 <- (1 - h)*sorted[k] + h*sorted[k - 1]

   c(x[1], q1, q2, sorted[n])
}

```{r, echo = T}
x <- c(0, 23, -1, 12, -10, -7, 1, -19, -6, 12, 1, -3)

idealf(x)
```

### 2.12

For the values: -1, -10, 2, 2, -7, -2, 3, 3, -6, 12, -1, -12, -6, 8, 6 compute the lower and upper quartiles (the ideal fourths).

```{r, echo = T}
x <- c(-1, -10, 2, 2, -7, -2, 3, 3, -6, 12, -1, -12, -6, 8, 6)

idealf(x)
```

### 2.13

Approximately how many values must be altered to make _q2_, the estimate of the upper quartile based on the ideal fourths, arbitrarily large?

_About $\frac{1}{4}$_

### 2.14

Argue that the smallest observed value, $X_1$, satisfies the definition of a measure of location.

_The smallest (or largest) value in a set defines the boundaries of the set. This is by definition a measure of location._

### 2.15

The height of 10 plants is measured in inches and found to be 12, 6, 15, 3, 12, 6, 21, 15, 18 and 12.

x <- c(12, 6, 15, 3, 12, 6, 21, 15, 18, 12)

xbar <- mean(x)

stopifnot( sum( x - xbar ) == 0 )

Verify that $\sum(X_i - \bar{X}) = 0$

### 2.16

For the data in the previous exercise, compute the range, variance and standard deviation.

n <- length(x)

range <- max(x) - min(x)
variance <- (1/(n-1))*sum( ( x - xbar )^2)
stdDev <- sqrt(variance)

$\bar{X} = `r xbar`$, Range = `r range`, $Var = `r variance`$, $\sigma = `r stdDev`$

\newpage

### 2.17

Use the rules of summation notation to show that it is always the case that $\sum(X_i - \bar{X}) = 0$.

$\bar{X} = \frac{1}{n}\sum{X_i}$

$\sum{X_i} - \frac{1}{n}\sum{X_i}$

$\ldots = \frac{1}{n}\sum{X_i} - \sum{X_i}$

$\ldots = \frac{1}{n}(\sum{X_i - X_i})$

$\ldots = \frac{1}{n}(0)$

$\ldots = 0$

### 2.18

Seven different thermometers were used to measure the temperature of a substance. The reading in degrees Celsius are -4.10, -4.13, -5.09, -4.08, -4.10, -4.09 and -4.12.

```{r, echo = T}
x <- c(-4.10, -4.13, -5.09, -4.08, -4.10, -4.09, -4.12)
n <- length(x)
xbar <- 1/n*sum(x)
variance <- (1/(n-1))*sum( (x - xbar)^2 )
stdDev <- sqrt(variance)
```

Find the variance and standard deviation.

$Var = `r variance`$, $\sigma = `r stdDev`$

### 2.19

A weightlifter's maximum bench press (in pounds) in each of 6 successive weaks was 280, 295, 275, 305, 300, 290.

Find the standard deviation.

```{r, echo = T}
x <- c(280, 295, 275, 305, 300, 290)
n <- length(x)
xbar <- 1/n*(sum(x))
variance <- (1/(n-1))*sum( (x - xbar)^2)
sqrt(variance)
```

### 2.20

For the values,

20, 121, 132, 123, 145, 151, 119, 133, 134, 130, 200

use the classic outlier detection rule to determine whether any outliers exist.

```{r, echo = T}
x <- c(20, 121, 132, 123, 145, 151, 119, 133, 134, 130, 200)

x[ abs(x - mean(x)) / sd(x) > 2 ]

### 2.21

Apply the boxplot rule and the MAD-median rule using the values in the preceding exercise. Note that the results differ, compared to using the classic rule.

x[ abs( x - median(x) ) / mad(x) > 2.27 ]

Explain why this happened.

_The classic method masks the value 200 as an outlier._

### 2.22

Consider the values,

0, 121, 132, 123, 145, 151, 119, 133, 134, 130, 250.

```{r, echo = T}
x <- c(0, 121, 132, 123, 145, 151, 119, 133, 134, 130, 250)

x[ ( abs(x - mean(x)) /sd(x) ) > 2 ]

Are the values 0 and 250 declared outliers using the classic outlier detection rule?

_Yes._

### 2.23

Verify that for the data in the previous exercise, the boxplot rule declares the values 0 and 250 outliers.

bounds <- quantile(x)[c(2,4)]

lower <- bounds[1] - 1.5*(bounds[2] - bounds[1])
upper <- bounds[2] + 1.5*(bounds[2] - bounds[1])

x[ x < lower | x > upper]

### 2.24

Consider the values

20, 121, 132, 123, 145, 151, 119, 133, 134, 240, 250

Verify that no outliers are found using the classic outlier detection rule.

```{r, echo = T}
x <- c(20, 121, 132, 123, 145, 151, 119, 133, 134, 240, 250)

stopifnot( length( x[ ( abs( x - mean(x) ) / sd(x) ) > 2] ) == 0 )

### 2.25

Verify that for the data in the previous exercise, the boxplot rule declares the values 20, 240 and 250 outliers.

```{r, echo = T}
x <- c(20, 121, 132, 123, 145, 151, 119, 133, 134, 240, 250)

bounds <- quantile(x)[c(2,4)]

lower <- bounds[1] - 1.5*(bounds[2] - bounds[1])
upper <- bounds[2] + 1.5*(bounds[2] - bounds[1])

x[ x < lower | x > upper]

### 2.26

What do the last three exercises suggest about the boxplot rule versus the classic rule for detecting outliers?

_The classic boxplot rule masks outliers._

### 2.27

What is the typical pulse rate (beats per minute) among adults? Imagine that you sample 21 adults, measure their pulse rate, and get:

80, 85, 81, 75, 77, 79, 74, 86, 79, 55, 82, 89, 73, 79, 83, 82, 88, 79, 77, 81, 82

```{r, echo = T}
x <- c(80, 85, 81, 75, 77, 79, 74, 86, 79, 55, 82, 89, 73, 79, 83, 82, 88, 79, 77, 81, 82)
```

compute the 20% trimmed mean.

n <- length(x)
o <- sort(x)
g <- floor(.2*n)
t <- o[ (g+1):(n-g) ]
tbar <- 1/(n - 2*g)*sum(t)

stopifnot(tbar == mean(x, trim = .2))
```

### 2.28

For the observations,

21, 36, 42, 24, 25, 36, 35, 49, 32

```{r, echo = T}
x <- c(21, 36, 42, 24, 25, 36, 35, 49, 32)
n <- length(x)

xbar <- 1/n * sum(x)

o <- sort(x)
g <- floor(.2*n)
t <- o[ (g+1):(n-g) ]
tbar <- 1/(n - 2*g)*sum(t)

M <- o[ (n + 1)/2 ]

Verify that the sample mean, 20% trimmed mean, and median are $\bar{X} = 33.33, \bar{X}_t = 32.9, and \ M = 35$

$\bar{X} = `r xbar`$, $\bar{X}_t = `r tbar`$, $M = `r M`$
